Highly Available Transactions: Virtues and Limitations

### 2013-03-26 writeup/motivation.tex
Another pass at writing

Peter Deutsch begins his classic list of ``Fallacies of Distributed
Computing'' with two concerns fundamental to distributed database
systems: ``\textit{1.)}  The network is reliable. \textit{2.)} Latency
is zero''~\cite{fallacies-deutsch}. On a single server, communication
systems are robust and relatively fast, and most single-node systems
do not need to prepare for partial component failure. However, these
assumptions are no longer valid in a distributed setting. To extend
Jim Gray's storage latency analogy, if memory is 1.5 hours from the
CPU, disk is 2 years away, especially across wide-area networks,
remote server accesses can exceed 30 years, with the added concern
that messages can be lost in transit or delayed indefinitely by
network partitions~\cite{gray-rules}. Accordingly, ignoring network
failures and latency ``cause[s] big trouble and painful learning
experiences''~\cite{fallacies-deutsch}.

In this section, we attempt to quantify the impact of network
partitions and latency in real-world deployments. While the importance
of network behavior is a topic of considerable speculation within the
database community~\cite{stonebraker2010errors}, there is mounting
evidence---both anecdotal and experimental---that partitions occur and
latency is non-negligible.

\subsection{Partitions}

Network partitions---or, informally, communication link failures
between participants in a distributed system---are challenging. A
system facing network partitions faces the impossibility of
simultaneously maintaining a correct ``global'' view of system state
while simultaneously providing operation to all of its
clients~\cite{davidson-survey}. For networks that do not partition,
maintaining semantic guarantees and available operation is
simple~\cite{stonebraker2010errors}. However, given the
\textit{possibility} of partitions, a system must be prepared to
either lose availability of operations or its ability to reason about
global state.

According to James Hamilton, Vice President and Distinguished Engineer
on the Amazon Web Service team, ``network partitions should be rare
but net gear continues to cause more issues than it
should''~\cite{hamilton-partitions}. Anecdotal evidence confirms
Hamilton's assertion. In April 2011, a network misconfiguration led to
a twelve-hour long series of outages across the Amazon EC2 and RDS
services~\cite{amazon-netpartition}. Subsequent misconfigurations and
partial failures such as another EC2 failure in October 2012 have led
to full site outages for web services like Reddit, Foursquare, and
Heroku~\cite{ec2-downsites}. At a global scale, hardware
failures---like the 2011 outages in Internet backbones in North
America and Europe due a router bug~\cite{juniper-partition}---and
misconfigurations---such as BGP faults in 2008~\cite{pakistan-youtube}
and 2010~\cite{research-experiment-partition}---can cause widespread
partitions. Many of our discussions with practitioners---especially
those operating on public cloud infrastructure---as well as reports
from large-scale operators like Google~\cite{dean-keynote} confirm
that partitions are a reality.

Several recent studies rigorously quantify partition behavior. A 2011
study of several Microsoft datacenters found a mean of 40.8 network
link failures per day (95th percentile: 136), with a median time to
repair of around five minutes, and up to one week. Perhaps
surprisingly, network redundancy only reduces the median impact of
failures by up to 40\%, meaning providers cannot easily reduce the
incidence of partitions~\cite{sigcomm-dc}. A 2010 study of over 200
wide-area routers found an average of 16.2--302.0 failures per link
per year with an average annual downtime of 24--497 minutes per link
per year (95th percentile at least 34 hours)~\cite{sigcomm-wan}. In
HP's managed enterprise networks, WAN, LAN, and connectivity problems
account for 28.1\% of all customer support tickets while 39\% of
tickets relate to network hardware.  The median incident duration for
highest priority tickets ranges from 114--188 minutes and up to a full
day for all tickets.~\cite{turner2012failure}. The incidence of
network failures is further confirmed by several prior studies showing
for example, on Sprint's WANs, a median time to repair between 2 and
1000 seconds, and a median time between failures of approximately 3000
seconds.~\cite{ip-backbone-failures} and frequent path routing
failures on the Internet~\cite{labovitz-failures}. Indeed, isolating,
quantifying, and designing for network failures is an area of active
research in networking community~\cite{surviving-failures-bodik,
  uw-failure-networks}.

@@

\subsection{Latency}

Even with fault-free networks, distributed systems face the challenge
of communication latency, Deutsch's second ``Fallacy.'' Fundamentally,
the speed at which two servers can communicate is (according to modern
physics) bounded by the speed of light. Two servers located on
opposite sides of the Earth face, in the best case---communicating via
a hypothetical link crossing through the center of the
planet---require a minimum 85.1 ms RTT (133.7 ms if sent at ground
level). As services are replicated to multiple, geographically
distinct sites, the cost of communication increases this upper bound.

In real deployments, latencies are higher than the speed of light due
to routing, congestion, and processing times; if not, servers a
kilometer apart would enjoy a 6.7 $\mu$s RTT. To illustrate the
difference between intra-datacenter, inter-datacenter, and
inter-planetary networks, we performed a measurement study of network
behavior on Amazon's EC2, a widely-used public compute cloud. We
measured one week of ping times between all seven EC2 regions, across
three ``availability zones'', or closely co-located datacenters, in
the Eastern United States, and within a single ``availabilty zone,''
at a granularity of 1s (dataset to be released).

We summarize the results of our network measurement study in
Table~\ref{table:rtt}. On average, inter-datacenter communication (1c)
is between 1.82 and 6.38 times faster than across geographically
co-located datacenters (1b) and 40 and 647 times faster than across
geographically distributed datacenters (1a). The cost of wide-area
communication exceeds the speed of light: for example, communicating
from S\~{a}o Paulo to Singapore is lower-bounded by speed-of-light
transit RTT of 106.7 ms but, on average, incurs a 362.8 ms RTT (95th
percentile: 649ms). As shown in Figure~\ref{fig:rtt}, the distribution
of latencies varies between links, but the trend---in line with other
recent studies~\cite{mdcc, redblue}---is clear: remote communication
has a substantial cost.

@@

\subsection{Modern Systems}
\label{sec:modernacid}

Mitigating the effects of partitions and network latency is a
difficult task, requiring system desigers to make hard trade-offs. In
this section, we briefly discuss known semantic limits for systems
wishing to maintain high availability and their impact on modern
database systems. While the gold standard of ACID databases is
unavailable, we show that many database systems operate under much
weaker models, motivating the study of these models in a highly
available context in the remainder of this paper.

Most notable among existing impossibility results for distributed
systems is Brewer's CAP Theorem, which, as formally proven by Gilbert
and Lynch, states that it is impossible to maintain linearizability
(``C,'' or ``consistency''---albeit poorly named), or the ability to
read the last completed write, and replica availability (``A'') in the
presence of network partitions (``P'')~\cite{gilbert-cap}. Brewer's
Theorem has roots in decades of distributed database
designs~\cite{davidson-survey} and has had a substantial impact on
recent large-scale distributed systems design. Yet, while CAP is
narrowly scoped to the impossibility linearizability, it is often
interpreted to preclude a larger set of guarantees, including ACID, or
transactional atomicity, consistency, isolation and durability
guarantees.

Database designers and researchers have long realized that
serializability---the gold standard of ACID isolation, which
guarantees application-level consistency---is not achievable in a
highly available system~\cite{davidson-survey}.  However, database
systems offer a range of ACID properties besides
serializability. Within a single-node database, the coordination
penalties associated with ensuring equivalence with a serial execution
can be severe and manifest themselves in the form of concurrency
control contention (and, subsequently, performance degradation,
scalability limitations, and, often, external aborts like
deadlocks)~\cite{gray-isolation}. Instead, databases offer a host of
so-called \textit{weak isolation} models that allow varying
restrictions on the space of schedules that are allowable by the
system~\cite{adya}. None of these weak isolation models guarantees
serializability, but the benefits of these models can outweigh the
costs of possible application-level consistency anomalies that arise
from their use.

To understand the prevalence of these weak isolation models, in an
earlier workshop paper~\cite{hat-hotos}, we surveyed the default and
maximum isolation models provided by 18 relational databases, often
claiming to provide ``ACID'' or ``NewSQL'' functionality. As shown in
Table~\ref{table:existing}, only three out of 18 databases provides
serializability by default, and at least eight do not provide
serializability as an option at all. This is particularly surprising
when we consider the widespread deployment of many of these databases,
like Oracle 11g, which are known to power major businesses and product
functionality. Given that these transactional models are frequently
used, our inability to provide serializability in arbitrary HATs is
non-fatal to practical applications. If application writers and
database vendors have already decided that the benefits of weak
isolation outweigh potential application inconsistencies, then, as we
will see, in a highly available environment, similar decisions may be
appropriate.

\begin{table}
\begin{small}
\begin{tabular}{|l|c|c|}
\hline
Database & Default & Maximum\\\hline
Actian Ingres 10.0/10S~\cite{actian-docs} & S & S\\
Aerospike~\cite{aerospike-docs} & RC & RC\\
Akiban Persistit~\cite{akiban-docs} & SI & SI\\
Clustrix CLX 4100~\cite{clustrix-docs} & RR & ?\\
Greenplum 4.1~\cite{greenplum-docs} & RC & S \\
IBM DB2 10 for z/OS~\cite{db2-docs} & CS & S\\
IBM Informix 11.50~\cite{informix-docs} & Depends & RR\\
MySQL 5.6~\cite{innodb-docs} & RR & S \\
MemSQL 1b~\cite{memsql-docs} & RC & RC\\
MS SQL Server 2012~\cite{ms-sql-docs} & RC & S \\
NuoDB~\cite{nuodb-docs} & CR & CR\\
Oracle 11g~\cite{oracle-docs} & RC & SI\\
Oracle Berkeley DB~\cite{bdb-reg-docs} & S & S\\
Oracle Berkeley DB JE~\cite{bdb-je-docs} & RR & S\\
Postgres 9.2.2~\cite{postgres-docs} & RC & S\\
SAP HANA~\cite{hana-docs} & RC & SI\\
ScaleDB 1.02~\cite{scaledb-docs} & RC & RC\\
VoltDB~\cite{voltdb-docs} & S & S\\
\hline
\multicolumn{3}{|p{7cm}|}{{\footnotesize RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}}\\\hline

\end{tabular}
\caption{Default and maximum isolation levels for ACID and NewSQL
  databases as of January 2013 (from
  \protect\cite{needed-hotos}).}\vspace{-1.5em}
\label{table:existing}
\end{small}
\end{table}

The primary challenge in providing weak isolation in a HAT context is
that, unfortunately, it is unclear \textit{which} of these isolation
guarantees can be provided with high availability. The existing
algorithms for providing weak isolation are typically presented for a
single-node context and are therefore unavailable. We are not aware of
any prior literature that provides guidance as to the relationship
between weak isolation and high availability. Prior work has examined
the relationship between serializability and high
availability~\cite{davidson-survey} and weak isolation on a
single-server~\cite{adya, ansicritique} but never weak isolation and
high availability taken together. We believe that this apparent gap in
the database literature is a possible contributing factor to the
often-apparent attitude that general-purpose transactional semantics
are either too expensive to provide in many modern distributed
databases---except under special circumstances, such as providing
operations over groups of co-located data items.

>>>>

Peter Deutsch begins his classic list of ``Fallacies of Distributed
Computing'' with two concerns fundamental to distributed database
systems: ``\textit{1.)}  The network is reliable. \textit{2.)} Latency
is zero''~\cite{fallacies-deutsch}. On a single server, communication
channels are robust and relatively fast, and few single-server systems
need to account for a possibility of partial system failure. However,
both of these assumptions are no longer valid in a distributed
setting. To extend Jim Gray's storage latency analogy, if memory is
1.5 hours from the CPU and disk is 2 years away, then, especially
across wide-area networks (WANs), remote server accesses can exceed 30
years, with the added concern that messages can be lost in transit or
delayed indefinitely by network
partitions~\cite{gray-rules}. Accordingly, ignoring network failures
and latency can ``cause big trouble and painful learning
experiences''~\cite{fallacies-deutsch}. As an example of an actual system rearchitecture,
less than one year after its announcement, Yahoo!'s PNUTS developers
explicitly added a support for weaker, highly available operation,
explaining that ``strict adherence [to strong consistency] leads to
difficult situations under network partitioning or server
failures...in many circumstances, applications need a relaxed
approach''~\cite{pnuts-update}.

In this section, we quantify the impact of network partitions and
latency in several real-world deployments. While the importance of
network behavior is a topic of considerable speculation within the
database community~\cite{stonebraker2010errors}, there is mounting
evidence---both anecdotal and experimental---that partitions occur and
network latencies are non-negligible.

\subsection{Partitions in the Real World}

Network partitions---or, informally, communication link failures
between participants in a distributed system---are challenging. A
system facing network partitions faces the impossibility of
simultaneously maintaining a correct ``global'' view of system state
while simultaneously providing operation to all of its
clients~\cite{davidson-survey}. For networks that do not partition,
maintaining semantic guarantees and available operation is
simple~\cite{stonebraker2010errors}. However, given the
\textit{possibility} of partitions, a system must be prepared to
either lose availability of operations or its ability to reason about
global state. In this section, we present evidence for the frequency
and non-triviality of WAN and LAN partitions.

According to James Hamilton, Vice President and Distinguished Engineer
on the Amazon Web Services team, ``network partitions should be rare
but net gear continues to cause more issues than it
should''~\cite{hamilton-partitions}. Anecdotal evidence confirms
Hamilton's assertion. In April 2011, a network misconfiguration led to
a twelve-hour long series of outages across the Amazon EC2 and RDS
services~\cite{amazon-netpartition}. Subsequent misconfigurations and
partial failures such as another EC2 outage in October 2012 have led
to full site disruptions for popular web services like Reddit,
Foursquare, and Heroku~\cite{ec2-downsites}. At a global scale,
hardware failures---like the 2011 outages in Internet backbones in
North America and Europe due a router
bug~\cite{juniper-partition}---and misconfigurations---such as BGP
faults in 2008~\cite{pakistan-youtube} and
2010~\cite{research-experiment-partition}---can cause widespread
partitions. Many of our discussions with practitioners---especially
those operating on public cloud infrastructure---as well as reports
from large-scale operators like Google~\cite{dean-keynote} confirm
that partitions are a reality for service operators today.

Several recent studies quantify partition behavior more rigorously. A
2011 study of several Microsoft datacenters found a mean of 40.8
network link failures per day (95th percentile: 136), with a median
time to repair of around five minutes (and up to one week). Perhaps
surprisingly, provisioning redundant networks only reduces impact of
failures by up to 40\%, meaning network providers cannot easily
curtail partition behavior~\cite{sigcomm-dc}. A 2010 study of over 200
wide-area routers found an average of 16.2--302.0 failures per link
per year with an average annual downtime of 24--497 minutes per link
per year (95th percentile at least 34 hours)~\cite{sigcomm-wan}. In
HP's managed enterprise networks, WAN, LAN, and connectivity problems
account for 28.1\% of all customer support tickets while 39\% of
tickets relate to network hardware.  The median incident duration for
highest priority tickets ranges from 114--188 minutes and up to a full
day for all tickets~\cite{turner2012failure}. The incidence of network
failures is further confirmed by several prior studies showing for
example, on Sprint's WANs, a median time to repair between 2 and 1000
seconds, and a median time between failures of approximately 3000
seconds~\cite{ip-backbone-failures}, as well as frequent path routing
failures on the Internet~\cite{labovitz-failures}. Isolating,
quantifying, and designing for these network failures is an area of active
research in networking community~\cite{surviving-failures-bodik,
  uw-failure-networks}.

@@

\subsection{Latency: An Omnipresent Concern}

Even with fault-free networks, distributed systems face the challenge
of communication latency, Deutsch's second ``Fallacy.'' In this
section, we quantify expected latencies, which are often
huge---hundreds of milliseconds---in a geo-replicated context.

Fundamentally, the speed at which two servers can communicate is
(according to modern physics) bounded by the speed of light. Two
servers located on opposite sides of the Earth face, in the best
case---communicating via a hypothetical link crossing through the
center of the planet---require a minimum 85.1 ms RTT (133.7 ms if sent
at ground level). As services are replicated to multiple,
geographically distinct sites, the cost of communication increases to
this upper bound.

In real deployments, latencies are actually higher than the speed of
light dictates due to routing, congestion, and processing times; if
not, servers a kilometer apart would enjoy a 6.7 $\mu$s RTT. To
illustrate the difference between intra-datacenter, inter-datacenter,
and inter-planetary networks, we performed a measurement study of
network behavior on Amazon's EC2, a widely-used public compute
cloud. We measured one week of ping times between all seven EC2
regions, across three ``availability zones'', or closely co-located
datacenters, and within a single ``availabilty zone,'' at a
granularity of 1s (dataset to be released).

We summarize the results of our network measurement study in
Table~\ref{table:rtt}. On average, inter-datacenter communication (1c)
is between 1.82 and 6.38 times faster than across geographically
co-located datacenters (1b) and between 40 and 647 times faster than
across geographically distributed datacenters (1a). The cost of
wide-area communication exceeds the speed of light: for example,
communicating from S\~{a}o Paulo to Singapore is lower-bounded by
speed-of-light transit RTT of 106.7 ms but, on average, incurs a 362.8
ms RTT (95th percentile: 649ms). As shown in Figure~\ref{fig:rtt}, the
distribution of latencies varies between links, but the trend---in
line with other recent studies~\cite{mdcc, redblue}---is clear: remote
communication has a substantial cost.

@@

\subsection{CAP, Isolation, and Modern Systems}
\label{sec:modernacid}

Mitigating the effects of partitions and network latency is a
difficult task, requiring systems designers to make fundamental
trade-offs. In this section, we briefly discuss known semantic limits
for systems wishing to maintain high availability and their impact on
modern database systems. While the gold standard of ACID databases is
unavailable, we show that many database systems operate under much
weaker models, motivating the study of these models in a highly
available context in the remainder of this paper.

One of the most popularly celebrated impossibility results for
distributed systems is Brewer's CAP Theorem, which, as formally proven
by Gilbert and Lynch, states that it is impossible to maintain
linearizability (``C,'' or ``consistency''---albeit poorly named), or
the ability to read the last completed write, and replica availability
(``A'') in the presence of network partitions
(``P'')~\cite{gilbert-cap}. Brewer's Theorem has roots in decades of
distributed database designs~\cite{davidson-survey} and has had a
substantial impact on recent large-scale distributed systems design:
while CAP is narrowly scoped to the impossibility of building highly
available linearizable systems linearizability, it is often
interpreted to preclude a larger set of guarantees, including ACID, or
transactional atomicity, consistency, isolation and durability
guarantees, as evidenced by the lack of highly available systems
providing them.  As an example of an production database redesign
explicitly attributable to partition behavior, less than one year
after its announcement, Yahoo!'s PNUTS system developers explicitly
added a support for weaker, highly available operation, explaining
that ``strict adherence [to strong consistency] leads to difficult
situations under network partitioning or server failures...in many
circumstances, applications need a relaxed
approach''~\cite{pnuts-update}.

Database designers and researchers have long realized that
serializability---the gold standard of ACID isolation, which
guarantees application-level consistency---is not achievable in a
highly available system~\cite{davidson-survey}.  However, database
systems offer a range of ACID properties besides
serializability. Within a single-node database, the coordination
penalties associated with ensuring equivalence with a serial execution
can be severe and manifest themselves in the form of concurrency
control contention (and, subsequently, performance degradation,
scalability limitations, and, often, external aborts like
deadlocks)~\cite{gray-isolation}. Instead, databases offer a host of
so-called \textit{weak isolation} models that allow varying
restrictions on the space of schedules that are allowable by the
system~\cite{adya}. None of these weak isolation models guarantees
serializability, but the benefits of these models can often outweigh
costs of possible consistency anomalies that might arise
from their use.

To understand the prevalence of these weak isolation models, we
recently surveyed, we surveyed the default and maximum isolation
models provided by 18 relational databases, often claiming to provide
``ACID'' or ``NewSQL'' functionality~\cite{hat-hotos}. As shown in
Table~\ref{table:existing}, only three out of 18 databases provides
serializability by default, and at least eight do not provide
serializability as an option at all. This is particularly surprising
when we consider the widespread deployment of many of these
non-serializable databases, like Oracle 11g, which are known to power
major businesses and product functionality. Given that these
transactional models are frequently used, our inability to provide
serializability in arbitrary HATs is non-fatal for practical
applications. If application writers and database vendors have already
decided that the benefits of weak isolation outweigh potential
application inconsistencies, then, in a highly available environment
that prevents serializability entirely, similar decisions may be
appropriate.

\begin{table}
\begin{small}
\begin{tabular}{|l|c|c|}
\hline
Database & Default & Maximum\\\hline
Actian Ingres 10.0/10S & S & S\\
Aerospike & RC & RC\\
Akiban Persistit & SI & SI\\
Clustrix CLX 4100 & RR & ?\\
Greenplum 4.1 & RC & S \\
IBM DB2 10 for z/OS & CS & S\\
IBM Informix 11.50 & Depends & RR\\
MySQL 5.6 & RR & S \\
MemSQL 1b & RC & RC\\
MS SQL Server 2012 & RC & S \\
NuoDB & CR & CR\\
Oracle 11g & RC & SI\\
Oracle Berkeley DB & S & S\\
Oracle Berkeley DB JE & RR & S\\
Postgres 9.2.2 & RC & S\\
SAP HANA & RC & SI\\
ScaleDB 1.02 & RC & RC\\
VoltDB & S & S\\
\hline
\multicolumn{3}{|p{7cm}|}{{\footnotesize RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}}\\\hline

\end{tabular}
\caption{Default and maximum isolation levels for ACID and NewSQL
  databases as of January 2013 (from
  \protect\cite{hat-hotos}).}\vspace{-1.5em}
\label{table:existing}
\end{small}
\end{table}

The primary challenge in providing weak isolation in a HAT context is
that, unfortunately, it is unclear \textit{which} of these isolation
guarantees can be provided with high availability. The existing
algorithms for providing weak isolation are typically designed for a
single-node context and are, to the best of our knowledge, unavailable
due to a reliance on concurrency control mechanisms like locking that
are not resilient to partial failure. Morever, we are not aware of any
prior literature that provides guidance as to the relationship between
weak isolation and high availability: prior work has examined the
relationship between serializability and high
availability~\cite{davidson-survey} and weak isolation on a
single-server~\cite{adya, ansicritique} but never weak isolation and
high availability taken together. We believe that this apparent gap in
the database literature is a possible contributing factor to the
often-apparent attitude that general-purpose transactional semantics
are either too expensive to provide in many modern distributed
databases---except under special circumstances, such as providing
operations over groups of co-located data items.
